Describe and Explore: Using feature-based time series clustering to identify meaningful structures in intensive longitudinal data

### 2023-01-08 manuscript.tex
:sparkles: nonlinearity

Fortunately, past conceptual and empirical efforts offer valuable discussions of common time series features in psychological research. While the final selection of features should always be driven by the research questions and field-specific conventions, we can build a practical toolbox of meaningful time series features for psychological data. In particular, we propose to focus on six features based on common psychological research questions and recent works on affect dynamics \citep[e.g.,][]{dejonckheere2019, kuppens2017, adya2001}. An overview of the proposed features, their substantive interpretations, and mathematical operationalizations is available in \tblref{tab:esmFeatures}.

To understand emotion dynamics, \citet{kuppens2017} originally proposed four features: (1) within-person variability, (2) co-variance or ICC, (3) inertia or autocorrelation, and (4) cross-lagged correlations. These features were then extended by \citet{krone2018}, adding (5) innovation variance, and (6) mean intensity. \citet{krone2018} even built a parametric model to tentatively cluster study participants. From a slightly different perspective \citet{dejonckheere2019} later added three additional features for psychological time series: (7) instability (8) interdependence (i.e., network density), and (9) diversity \citep[i.e., Gini coefficient; also see][]{wendt2020}\footnote{It should be noted that also within the psychological literature alternative summaries have been proposed that for example include measurement distribution, nonlinear developments, or categorical states. As an example, \citet{kiwuwa-muyingo2011} proposed to extract clinically meaningful states for medical adherence data and suggests these states as meaningful time series features.}.

This selection of the proposed six time-series features is in no way exhaustive or imperative. Both using a purely data-driven approach or selecting other aspects to summarize the time series are legitimate options. The list seeks to offer a practical toolbox of features that are common and meaningful to psychological research questions and -practice (see \tblref{tab:esmFeatures} for a more complete overview).

@@

\paragraph{Nonlinearity} Changes in psychology are not always linear \citep{hayes2007}

Nonlinear trends are important in two regards. Firstly, nonlinearity is a deviation from linear trends and secondly, the shape and structure of the nonlinear trend.

``Some examples of nonlinear DEs have already been proposed in studies of ovulatory regulation (Boker, Neale, \& Klump, 2014), circadian rhythms (E. N. Brown \& Luithardt, 1999), cerebral development (Thatcher, 1998), substance use (Boker & Graham, 1998), cognitive aging (Chow \& Nesselroade, 2004), parent-child interactions (Thomas \& Martin, 1976), dyadic relationships (Chow, Ferrer, \& Nesselroade, 2007); and sudden transitions in attitudes (van der Maas, Kolstein, \& van der Pligt, 2003)." \citep{chow2019}

Questions like: ``Is the development of anxiety a nonlinear process?'' are mathematically captured using nonlinearity parameters such as the bicoherence metrics \citep{cuddy2009}.

The structure of nonlinear trends is often mathematically more difficult to capture \citep{castro-alvarez2022a}. One common way is to capture the trends using polynomial coefficients or more broadly by how `wiggly' the line is (e.g., estimated degrees of freedom of GAM spline models; similar to the number of spikes \citealp[]{caro-martin2018}).

depression most likely non-linear  \citep{hosenfeld2015}. Patients in recovery showed sudden changes in the improvement of depression \citep{helmich2020a}

>>>>

Fortunately, past conceptual and empirical efforts offer valuable discussions of common time series features in psychological research. To understand emotion dynamics, \citet{kuppens2017} originally proposed four features: (1) within-person variability, (2) co-variance or ICC, (3) inertia or autocorrelation, and (4) cross-lagged correlations. These features were then extended by \citet{krone2018}, adding (5) innovation variance, and (6) mean intensity. \citet{krone2018} even built a parametric model to tentatively cluster study participants. From a slightly different perspective \citet{dejonckheere2019} later added three additional features for psychological time series: (7) instability (8) interdependence (i.e., network density), and (9) diversity \citep[i.e., Gini coefficient; also see][]{wendt2020}\footnote{It should be noted that also within the psychological literature alternative summaries have been proposed that for example include measurement distribution, nonlinear developments, or categorical states. As an example, \citet{kiwuwa-muyingo2011} proposed to extract clinically meaningful states for medical adherence data and suggests these states as meaningful time series features.}.

Some of the proposed features are not necessarily well suited to summarize time series for feature-based clustering and some key conceptual features are not well represented in the emotion dynamics literature, which has been so formative for the analysis of psychological time series data. In particular, covariances and cross-lagged correlations often produce a staggering amount of parameters and might not necessarily summarize the existing data enough \citep{ernst2021}. Others such as network density parameters, used to summarize variable interdependence, might not always be meaningful for psychological data \citep{bringmann2019a}. At the same time, linear and nonlinear trends are not captured by the features commonly proposed for psychological time series, because the features are often developed for stationary models \citep[e.g.,][]{krone2018}.

Thus, while the final selection of features should always be driven by the research questions and field-specific conventions, we hope to build a practical toolbox of meaningful time series features for psychological data. In particular, we propose to focus on six features based on common psychological research questions and recent works within the field: (1) central tendency, (2) variability, (3) instability, (4) self-similarity, (5) linear trend, and (6) nonlinearity. We offer a short introduction of each of these six features below as a practical starting point for researchers and practitioners interested in feature-base clustering of psychological time series data. An overview of the proposed features, their substantive interpretations, and mathematical operationalizations is also available in \tblref{tab:esmFeatures}.

@@

\paragraph{Nonlinearity} Changes in psychology are not always linear, instead nonlinearity is is a common feature of psychological time series \citep{hayes2007}. As an example, episodic disorders, such as depression, are most likely best described as non-linear systems \citep{hosenfeld2015}. Similarly, patients in recovery of depression showed sudden changes in the improvement of depression \citep{helmich2020a}. But also substance abuse \citep{boker1998} or attitude changes rarely develop linearly \citep{vandermaas2003}. For psychological researchers, nonlinear developments are, thus, important in two regards. Firstly, nonlinearity as a deviation from linear trends is an important feature defining many psychological time series and secondly, the shape and structure of the nonlinear trend can be essential in understanding the important changes and transitions of human experiences \citep{chow2019}. Conceptually, researchers might accordingly have research question about the type of the development: ``Is the development of anxiety a nonlinear process?'' as well as the shape and structure of the development: ``Does anxiety change as a sudden or smooth transition?'' Accordingly, researchers can capture the type of development with nonlinearity parameters such as the bicoherence metrics \citep{cuddy2009}. The structure of nonlinear trends is often mathematically more difficult to capture \citep{castro-alvarez2022a}. Common ways to capture the trends include the use of polynomial- and differential equation parameters or more broadly to capture how `wiggly' the line is (e.g., estimated degrees of freedom of GAM spline models; similar to the number of spikes \citealp[]{caro-martin2018}).

It should be noted that this selection of the proposed six time-series features is in no way exhaustive or imperative. Both using a purely data-driven approach or selecting other aspects to summarize the time series are legitimate options. The list seeks to offer a practical toolbox of features that are common and meaningful to psychological research questions and -practice but are also easy to extract and summarize a broad range of developments without asserting strict assumptions (see \tblref{tab:esmFeatures} for a more complete overview).
